Missing The Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-11-18 Manuscript.Rmd
Add reference to intro

Iterative imputation has become the de facto approach to accommodate the ubiquitous problem of missing data. While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence. Since there is no consensus on how to evaluate the convergence properties of iterative imputation algorithms, our study provides insight into identifying non-convergence. We found that--in the cases considered--inferential validity was achieved after five to ten iterations, much earlier than indicated by diagnostic methods. We conclude that it never hurts to iterate longer, but such calculations hardly bring added value.

@@

Missing data pose a ubiquitous threat to anyone who aims to obtain unbiased, confidence-valid statistical inferences. The de facto standard to accommodate missing data is to 'impute' (i.e., fill in) any missing values in an incomplete dataset. To generate imputed values, most imputation software packages rely on iterative algorithms. [Why iterative? We estimate the posterior predictive distribution of the missing values; our estimates become better as we go; there should be a point at which our estimates are non-improving.] Iterative imputation techniques have proven to be powerful tools to draw valid inference under many missing data circumstances [@rubin87; @buur18].

>>>>

Iterative imputation has become the de facto standard approach to accommodate the ubiquitous problem of missing data. While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence. Our study provides insight into identifying non-convergence in iterative imputation algorithms. We show that these algorithms can yield correct outcomes even when a converged state has not yet formally been reached. In the cases considered, inferential validity is achieved after five to ten iterations, much earlier than indicated by diagnostic methods. We conclude that it never hurts to iterate longer, but such calculations hardly bring added value.

@@

<!-- Iterative imputation has become the de facto standard approach to accommodate missing data problems. With iterative imputation, we 'impute' (i.e., fill in) plausible values in an incomplete dataset. These plausible values are generated using algorithms such as 'MICE' [@mice]. These algorithms sequence through the variables in an incomplete dataset  -->

<!-- Iterative imputation entails using algorithms such as MICE [@mice] to replace missing values by plausible values. These plausible values are sampled from the distribution of each missing entry, and draw a plausible value from this distribution to replace the missing value by. -->

<!-- While it is widely accepted that the technique can yield valid inferences under many different missing data circumstances, these inferences all rely on algorithmic convergence. We fill in the missing data by modeling a multivariate distribution for each missing entry. We update dis model by sequencing through the variables See Figure 1. Van Buuren et al. (2006) have shown that just a few iterations maybe sufficient. Most imputation software packages use a default value of five or ten iterations. But how can we know did this is sufficient?  -->
<!-- There is currently no consensus on how to evaluate the convergence properties of iterative imputation algorithms. Imputation algorithms such as mice are a type of Gibbs sampler. It would does make sense to evaluate the convergence of iterative imputation algorithms using common MSMC non convergence identifiers. This has not been done.  -->
<!-- With MCMC algorithms there is not a single point at which convergence is reached.  -->

Iterative imputation has become the de facto standard approach to accommodate missing data. With iterative imputation, we 'impute' (i.e., fill in) plausible data values in the place of missing values in an incomplete dataset. We obtain these plausible values by sampling from a distribution that we model for each missing entry and update sequentially. @buur06 have shown that with an iterative imputation algorithm such as 'MICE', just a few iterations are enough to yield unbiased, confidence-valid statistical inferences.

<!-- Missing data pose a ubiquitous threat to anyone who aims to obtain unbiased, confidence-valid statistical inferences. The de facto approach to accommodate missing data is to 'impute' (i.e., fill in) any missing entries in an incomplete data. With iterative imputation, we model the distribution of each missing entry, and draw one or more plausible values from it.  -->

<!-- With iterative imputation, we 'impute' (i.e., fill in) the missing data in an incomplete dataset by drawing values from the distribution of each missing entry. To generate imputed values, most imputation software packages rely on iterative algorithms. [Why iterative? We estimate the posterior predictive distribution of the missing values; our estimates become better as we go; there should be a point at which our estimates are non-improving.]  -->

Iterative imputation techniques have proven to be powerful tools to draw valid inference under many missing data circumstances [@rubin87; @buur18].
